---
output:
  md_document:
    variant: gfm
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      fig.path = "README_files/", dpi = 96)
```

# swehist

Boundaries of Swedish administrative units 1600-1990, as `sf` objects:
parishes, counties, municipalities, pastorships, deaneries, dioceses,
hundreds, magistrates courts, district courts, courts of appeal and
bailiwicks. Each unit is stored once per period in which its territory was
unchanged, with its predecessors and successors and the units it belonged
to. The package gives maps for any year or period, and matches parish and
unit names and codes in historical records to the maps.

The boundaries come from the historical GIS data of Riksarkivet (the Swedish
National Archives).

## Installation

```r
install.packages("swehist",
                 repos = c("https://junkka.r-universe.dev", "https://cloud.r-project.org"))
# or from GitHub
# remotes::install_github("junkka/swehist")
```

## Example

```{r example, fig.width = 6, fig.height = 8}
library(swehist)
library(ggplot2)

parishes <- get_boundaries(1900, "parish")
counties <- get_borders(1900, "county")

ggplot() +
  geom_sf(data = parishes, fill = "white", colour = "grey75", linewidth = 0.05) +
  geom_sf(data = counties, colour = "black", linewidth = 0.4) +
  theme_void()
```

Records that name a parish or give its code are linked to the map with
`match_units()`, and `assign_parent()` gives the county, pastorship, court
or other unit a parish belonged to in a given year:

```{r match}
m <- match_units(c("Lövånger", "Burträsk", "Nysätra"), county = "AC", date = 1880)
m[, c("input", "geom_id", "geom_name", "start", "end")]

assign_parent(m$geom_id, 1880, "pastorship")
```

`vignette("usage", package = "swehist")` shows the main tasks: maps for a
year or a period, linking records, parent and child units, unit histories.

## Datasets

```{r datasets, echo = FALSE, results = 'asis'}
d <- c(boundaries = 'Units of 11 types, one row per unit and period (sf)',
       relations = 'Successors between units, and transfers of territory',
       hierarchy = 'Which unit belonged to which, by year, with the evidence behind it',
       events = 'Creations, abolitions, renamings, mergers, splits and transfers',
       parish_link = 'Parish codes and names (Riksarkivet, SCB, Skatteverket)',
       parish_registry = "All parishes in Skatteverket's register, incl. those without territory",
       parish_meta = 'Parish codes and county',
       parish_codes = 'Every dated code (DDB, SCB, NAD) of every parish version',
       unit_variants = 'Name variants used by `match_units()`',
       quality_table = 'Coverage and evidence grade, by type and 50-year period',
       known_issues = 'Researched deviations that no source lets us correct',
       hist_town = 'County towns (sf)',
       sweden = 'National outline: the union of all parishes (sf)')
n <- vapply(names(d), function(x) nrow(get(x, envir = asNamespace('swehist'))), integer(1))
cat('| Dataset | Rows | Contents |', '|---|---:|---|',
    sprintf('| `%s` | %s | %s |', names(d), formatC(n, big.mark = ',', format = 'd'), d),
    sep = '\n')
```

All spatial data are in SWEREF99 TM (EPSG:3006).

## Administrative types

```{r types, echo = FALSE, results = 'asis'}
b <- sf::st_drop_geometry(swehist::boundaries)
sw <- c(parish = 'Socken, församling', county = 'Län', municipality = 'Kommun',
        pastorship = 'Pastorat', contract = 'Kontrakt', diocese = 'Stift',
        hundred = 'Härad, tingslag', magistrates_court = 'Domsaga, rådhusrätt',
        district_court = 'Tingsrätt', court_of_appeal = 'Hovrätt', bailiwick = 'Fögderi')
en <- c('Parish', 'County', 'Municipality', 'Pastorship', 'Deanery', 'Diocese', 'Hundred',
        'Magistrates court', 'District court', 'Court of appeal', 'Bailiwick')
r <- do.call(rbind, lapply(seq_along(sw), function(i){
  x <- b[b$type_id == names(sw)[i], ]
  data.frame(en = en[i], sw = sw[[i]], id = names(sw)[i],
             units = length(unique(x$unit_id)), versions = nrow(x),
             years = paste0(min(x$start), '-', max(x$end)))
}))
cat('| Type | Swedish | `type_id` | Units | Versions | Years |', '|---|---|---|---:|---:|---|',
    sprintf('| %s | %s | `%s` | %s | %s | %s |', r$en, r$sw, r$id,
            formatC(r$units, big.mark = ',', format = 'd'),
            formatC(r$versions, big.mark = ',', format = 'd'), r$years),
    sep = '\n')
```

A *unit* is the thing itself; a *version* is one period in which its territory did not
change. Regiments have no boundaries; `hierarchy` links them to their parishes.

## Sources

- **Riksarkivet** (Swedish National Archives): *Historiska GIS-kartor
  (information om territoriella indelningar i Sverige från 1500-talets slut
  till 1900-talets slut)* and the topographic register of the Nationell
  Arkivdatabas (NAD), released under CC0.
- **Skatteverket**: *Sveriges församlingar genom tiderna* (1989), for parish
  identifiers, names and the parish histories used to date pastorship and
  county links.
- **Statistics Sweden (SCB)**: parish codes.
- **The Demographic Data Base (DDB)**, Umeå University: parish names.

`data-raw/` has the build pipeline. It needs the Riksarkivet source files,
which are not in this repository (see `data-raw/README.md`).
`VALIDATION.md` summarises the checks of the data against independent
sources.

## Citation

`citation("swehist")` gives the reference.

## History

swehist grew out of my histmaps package (2015-2019,
<https://github.com/junkka/histmaps>), which I rebuilt as swehist in 2026 with
AI agents (Claude) doing much of the work.

**How the data were made.** I designed the data model, set the rules for which
sources count as evidence and in what order, and decided the cases where the
sources disagree. Within that design I used AI agents to write the build code
and to research the 1,310 correction rows that decide identity, membership and
dates, each row carrying the printed or online source it rests on
(`data-raw/model/corrections*.csv`). The *Sveriges statskalender* tables for
the courts and the deaneries were parsed from scanned text by scripts the agents
wrote, and a large language model read the parish histories in *Sveriges
församlingar genom tiderna* into dated tables (`data-raw/sfgt/` holds the
prompts and the output). The Skåne coordinate correction was fitted, and the
missing exclaves found, the same way. The sources are cited row by row and can
be checked, and the build is reproducible from them.

To measure accuracy, agents researched two held-out samples from printed
sources, without sight of the database. I checked about 100 of these facts by
hand, chosen by browsing the samples rather than by a fixed rule and including
many of the facts where the database and the source disagree. For each I
confirmed that the cited source, at the recorded place, gives the recorded unit
and that the unit is historically right, and for each disagreement also that
the database, not the fact, is wrong. I found a few minor faults and no reason
to doubt the facts as a whole. However, since the sample was not drawn at
random, it supports the facts without giving an error rate for them.

I have not checked the correction rows one by one. Treat a citation as the
warrant for a value, and check it yourself where a result matters to your work.

## License

The code is licensed MIT and the data CC BY 4.0; see `LICENSE.md`.
