---
title: "Using swehist"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using swehist}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  dpi = 72,
  message = FALSE
)
```

swehist holds the boundaries of Swedish administrative units 1600-1990:
parishes, counties, municipalities, pastorships, deaneries (contracts),
dioceses, hundreds, magistrates courts, district courts, courts of appeal
and bailiwicks. Each unit is stored once per period in which its territory
was unchanged, with the years it existed, the units it came from and went
on to, and the units it belonged to. This vignette shows the common tasks:
maps for a year, maps for a period, linking records by unit name or code,
and finding the parent units of a parish. The sources and how the data were
built are described in the dataset help pages (`?boundaries`,
`?hierarchy`, `?relations`) and in the data paper (`citation("swehist")`).

```{r load}
library(swehist)
library(sf)
library(dplyr)
library(ggplot2)
```

All maps are `sf` objects in SWEREF99 TM (EPSG:3006).

## A map for one year

`get_boundaries()` takes a year or a date and a type:

```{r counties}
counties <- get_boundaries(1900, "county")
counties
```

```{r counties_plot, fig.height = 7}
ggplot() +
  geom_sf(data = counties, fill = "grey95", colour = "grey40") +
  geom_sf(data = county_towns(1900), colour = "firebrick") +
  theme_void()
```

The number of units of each type in 1900:

```{r types}
types <- c("parish", "municipality", "county", "pastorship", "contract",
           "diocese", "hundred", "magistrates_court", "court_of_appeal",
           "bailiwick")
tibble(type = types,
       units = sapply(types, function(t) nrow(get_boundaries(1900, t))))
```

District courts exist from 1971 only. Not every type covers all of Sweden in
every year: hundreds, for instance, cover only southern and central Sweden,
because Norrland had tingslag instead. `get_boundaries()` warns when a map
covers less than 90% of the country; `?boundaries` lists the coverage of
each type.

## A map for a period

Parishes split and merged, so a parish map for 1850 does not fit records from
1900. Give a range of years to get a **period map**: units that changed
during the range are merged, so every unit on the map existed, with the same
territory, for the whole period.

```{r period}
pm <- get_boundaries(c(1850, 1900), "parish")
nrow(get_boundaries(1850, "parish"))
nrow(get_boundaries(1900, "parish"))
nrow(pm$map)
```

`pm$map` holds the merged units (`geomid`, with the merged `geom_ids`), and
`pm$lookup` maps each `geom_id` to its `geomid`. Use the lookup to aggregate
data recorded on the units of different years:

```{r period_lookup}
head(pm$lookup)
```

For example, with a count of events per parish and year (`geom_id`, `year`,
`n`), `left_join(events, pm$lookup, by = "geom_id")` followed by
`count(geomid, wt = n)` gives the counts on the period map.

## Linking records to units

Records usually name a parish or give a parish code. `match_units()` returns
the `geom_id` of the unit. A county (code, letter or name) narrows the search
when a name occurs in several counties; there is a Nysätra in Uppsala län
too:

```{r match}
match_units(c("Lövånger", "Burträsk", "Nysätra"), county = "AC")
```

Numbers are read as codes (forkod, nadkod, dedik or pid, from the number of
digits). Names can be in pre-1906 spelling:

```{r match_code}
match_units(248201)
match_units(c("Hvetlanda", "Carlskrona"))
```

`match_type` says how each match was made (by code, exact name, normalised
name or, with `fuzzy = TRUE`, edit distance). `multiple` is `TRUE` when more
than one unit fitted.

A parish name can refer to several versions of the parish. Give `date` to get
the version that existed then; `active_at_date` tells whether it did:

```{r match_date}
match_units(c("Byske", "Jukkasjärvi"), date = 1900)
```

`match_units()` also matches the other types (`type = "county"`,
`"municipality"` and so on). For parishes it uses `parish_registry`, which
includes about 840 parishes without territory of their own (bruks-,
garrison and hospital parishes); where their parent parish is known they
get its `geom_id`.

## Parent units

`assign_parent()` gives the unit of another type that each unit belonged to
at a date, following the links in `hierarchy`:

```{r parent}
ids <- match_units(c("Lövånger", "Burträsk", "Nysätra"), county = "AC",
                  date = 1880)$geom_id
assign_parent(ids, 1880, "county")
assign_parent(ids, 1880, "pastorship")
assign_parent(ids, 1880, "magistrates_court")
```

`get_children()` goes the other way: the units of one type inside a parent
unit.

```{r children, fig.height = 4}
malmohus <- get_children(1880, "county", "Malmöhus", "parish")
ggplot(malmohus) +
  geom_sf(fill = "grey95", colour = "grey40", linewidth = 0.2) +
  theme_void()
```

Regiments have no boundaries of their own, but their parishes do:

```{r regiment}
uppland <- get_children(1800, "regiment", "Uppland", "parish")
nrow(uppland)
```

## A unit's history

`unit_history()` follows `relations` backwards and forwards from a unit and
shows each version with the event that created it (origin, split, merge,
continuation):

```{r history}
unit_history("Byske", "parish", direction = "backward") %>%
  select(name, start, end, event)
```

## Borders

`get_borders()` turns polygons into border lines, for drawing one level on
top of another:

```{r borders, fig.height = 7}
ggplot() +
  geom_sf(data = get_boundaries(1900, "parish"),
          fill = "white", colour = "grey80", linewidth = 0.05) +
  geom_sf(data = get_borders(1900, "county"), colour = "black") +
  theme_void()
```

## Moving data between divisions

`crosswalk()` gives areal weights between two divisions, which is what you need
to carry a count from the parishes of one year to the units of another:

```{r crosswalk}
w <- crosswalk(c(1880, "parish"), c(1990, "municipality"))
w %>% filter(grepl("^Byske", from_name)) %>% select(to_name, km2, share_of_from)
```

`share_of_from` is the share of the source unit in each overlap, so a count
spread evenly over the parish is apportioned with it. `locate_point()` says
which unit held a place at a date, and `neighbours()` lists the units that
share a border:

```{r locate}
locate_point(c(18.07, 13.19), c(59.33, 55.71), 1880, "parish")
neighbours(1900, "county") %>% filter(a_name == "Uppsala län")
```

## What a unit was called, and what it rests on

Names, codes and memberships are dated, so a version carries the name that was
in force. `events` records the changes themselves:

```{r events}
data(events)
events %>% filter(grepl("Kyrkefalla|Tibro", detail), event == "renamed") %>%
  select(date, event, detail)
```

The levels are not equally well founded, and `quality_table` says what each
rests on: how much of the country a type covers in a period, and whether its
memberships come from a dated source, from a chain through another unit, from
an undated register link, or from geometry alone.

```{r quality}
data(quality_table)
quality_table %>% filter(period == "1850-1899") %>%
  select(parent_type, coverage, main_grade, sourced_or_better)
```

`known_issues` lists the boundary problems that research established but that
no source lets us correct, so you can see them rather than discover them:

```{r issues}
data(known_issues)
known_issues %>% filter(grepl("Malingsbo", name)) %>% select(name, start, end, issue)
```

## Things to know

- **Parishes before and after 1862.** Before the local government reform of
  1862 a parish (*socken*) was both a civil and a church unit. After it, the
  parish (*församling*) is a church unit and the municipality (*kommun*) the
  civil one. Parish boundaries continue across 1862, but their meaning
  changes.
- **Magistrates courts** include both rural courts (*domsagor*) and town
  courts (*rådhusrätter*), and the Hundred type holds the towns that stood
  outside the härad. The `kind` column tells them apart, as it does *städer*,
  *köpingar* and *landskommuner* among the municipalities.
- **Where the links come from.** Every link in `hierarchy` carries its
  `source` and a `grade`: a correction, a dated source (SCB, SFGT, *Sveriges
  statskalender*), a dated or undated register link, a chain through another
  unit, or spatial containment. `quality_table` sums that up by type and
  period.
- **Every unit is built from its parishes**, so the levels fit each other.
  Where no evidence established a unit's members it keeps the polygon
  Riksarkivet recorded for it, and `geometry_from` says so.
- **Regiments** were recruited from farms, and farms in one parish could
  serve different regiments. The regiment of a parish is therefore
  approximate.
- **Municipalities** in 1952 and 1971 miss some towns and market towns
  (*städer*, *köpingar*) that the source does not have.
- **The national outline** (`sweden`) is the union of all parishes. Large
  lakes that no parish includes (Vänern, Vättern) lie outside it.
- The data end in 1990.
